```{r packages}
library(GGally)
library(janitor)
library(corrplot)

library(ggfortify) ## PCA Plots
library(glmnet) ## Penalized Regression
library(MASS) ## Step Selection
library(MuMIn) ## Compare Model Performance
library(caret)
library(splines)
library(tidyverse)

set.seed(123)
```

```{r load-data}
house_prices <- read_csv("./cook_county_train_val.csv")

house_prices <- house_prices %>% clean_names()

# Drop meaningless variables
house_prices <- house_prices %>% select(-x1, -pin, -description)

# These columns only have one level per factor, so are not useful for modeling
house_prices <- house_prices %>%
  select(-use, -modeling_group)

# These are duplicates or modifications of other columns
house_prices <- house_prices %>%
  select(-neigborhood_code_mapping, -town_code, -neighborhood_code)

# Documentation says other improvements is not clean enough to be useful
house_prices <- house_prices %>%
  select(-other_improvements)

# Clean sale prices = 1
house_prices <- house_prices %>%
  filter(sale_price > 1)

# Codebook specifies that houses with missing age are set to 10.  Since we can't tell houses with missing age vs houses with true age 10, we decided to drop data points with age == 10.
house_prices <- house_prices %>%
  filter(age != 10)

# land_square_feet and lot_size are the same so we drop one
house_prices <- house_prices %>%
  select(-land_square_feet)

# there are lots of different fields treating date differently, for example sale month of year, sale half year, etc.  To cut down on redundant predictors, we are keeping sale year and sale month of year, as these fields contain the most information
house_prices <- house_prices %>%
  select(-sale_quarter, -sale_half_year, -sale_quarter_of_year, -sale_half_of_year)

# deed number is a unique ID that appears randomly assigned
house_prices <- house_prices %>%
  select(-deed_no)

# age_decade is just age / 10
house_prices <- house_prices %>%
  select(-age_decade)

# lots of zeroes for estimate_land and estimate_building.  We are interested in how good of a predictor of sale prices these estimates are, so we are dropping observations where these are 0
house_prices <- house_prices %>%
  filter(estimate_land != 0, estimate_building != 0)


# apartments ranges from 0 to 6, but there are no observations with 1 apartment.  We want to treat this variable as continuous, so we are transforming apartments = 0 to 1.  This means that an observation with 1 apartment is a free standing dwelling, 2+ apartments is an apartment complex.
house_prices <- house_prices %>%
  mutate(apartments = if_else(apartments == 0, 1, apartments))


house_prices <- house_prices %>%
  mutate_at(vars(property_class,
                 wall_material,
                 roof_material,
                 basement,
                 basement_finish,
                 central_heating,
                 other_heating,
                 central_air,
                 attic_type,
                 attic_finish,
                 design_plan,
                 cathedral_ceiling,
                 construction_quality,
                 site_desirability,
                 garage_1_material,
                 garage_1_attachment,
                 garage_1_area,
                 garage_2_material,
                 garage_2_attachment,
                 garage_2_area,
                 porch,
                 repair_condition,
                 multi_code,
                 census_tract,
                 multi_property_indicator,
                 o_hare_noise,
                 floodplain,
                 road_proximity,
                 pure_market_filter,
                 garage_indicator,
                 town_and_neighborhood,
                 sale_month_of_year),
            factor)

input_matrix <- model.matrix(sale_price ~ -1 + ., data = house_prices)
```

```{r eda, cache = TRUE}
for (var in c("building_square_feet", "lot_size", "estimate_land", "estimate_building", "age")) {
  print(var)
  plot <- ggplot(house_prices, aes(x = .data[[var]], y = log(sale_price))) +
    geom_point() +
    ggtitle(paste(var, "vs log Sale Price")) 
  print(plot)
}

ggplot(house_prices, aes(x = town_and_neighborhood, y = log(sale_price))) +
  geom_boxplot() +
  ggtitle("Town and Neighborhood vs log Sale Price")

ggplot(house_prices, aes(x = property_class, y = log(sale_price))) +
  geom_boxplot() +
  ggtitle("Property Class vs log Sale Price")

M <- cor(select_if(house_prices, is.numeric))
corrplot(M)

summary(house_prices) %>%
  knitr::kable()
```


```{r linear-models}

baseline <- lm(sale_price ~ building_square_feet, data = house_prices)

summary(baseline)

null_model <- lm(sale_price ~ 1, data = house_prices)
```

```{r full_model, eval = F}
full_model <- lm(sale_price ~ ., data = house_prices)
saveRDS(full_model, "full_model.RDS")
```

```{r full_model2}
full_model <- readRDS("full_model.RDS")
summary(full_model)
```

# Subset Selection
``` {r subset-selection, eval = F}
## Code below used to generate forward step fit
step_fit <- stepAIC(null_model,  scope = list(lower = null_model, upper = full_model), direction = "forward", steps = 5)
saveRDS(step_fit, "step_fit.RDS")
```

``` {r forward-selection-load}
## Load cached step fit model
step_fit <- readRDS("step_fit.RDS")

summary(step_fit)
```

# L_1/L_2 Regularized Regression

``` {r ridge-lasso, cache = TRUE}
lasso.cv <- cv.glmnet(x = input_matrix,
                      y = house_prices$sale_price,
                      alpha = 1,
                      nfolds = 5,
                      parallel = TRUE)

plot(lasso.cv)

plot(lasso.cv$glmnet.fit)

coef(lasso.cv, s = "lambda.min")

## Lambda min
lasso.cv$lambda.min

ridge.cv <- cv.glmnet(x = input_matrix,
                      y = house_prices$sale_price,
                      alpha = 0,
                      nfolds = 5,
                      parallel = TRUE)

plot(ridge.cv)

plot(ridge.cv$glmnet.fit)

coef(ridge.cv, s = "lambda.min")

## Lambda min
ridge.cv$lambda.min
```


# PCA

``` {r pca, cache = TRUE, eval = F}
pca <- prcomp(input_matrix)

plot(pca)

autoplot(pca,
         data = house_prices,
         color = "property_class",
         loadings = TRUE,
         loadings.color = "blue",
         loadings.label = TRUE,
         loadings.label.size = 5)

components <- pca[["x"]]

components <- data.frame(components)

components$PC3 <- -components$PC3

components$PC2 <- -components$PC2
```

```{r non-linear}
ggplot(house_prices, aes(x = building_square_feet, y = sale_price)) +
  geom_point()

### Generated with Claude 3.5 Sonnet
cv_natural_spline <- function(data, response, predictor, df = 3, k = 5) {
  # Create fold indices
  set.seed(123)  # for reproducibility
  folds <- createFolds(data[[response]], k = k, list = TRUE)
  
  # Store MSE for each fold
  mse_values <- numeric(k)
  
  # Perform k-fold CV
  for (i in seq_along(folds)) {
    # Split data into training and test
    test_indices <- folds[[i]]
    train_data <- data[-test_indices, ]
    test_data <- data[test_indices, ]
    
    # Fit natural spline on training data
    formula <- as.formula(paste(response, "~ ns(", predictor, ", df =", df, ")"))
    model <- lm(formula, data = train_data)
    
    # Make predictions on test data
    predictions <- predict(model, newdata = test_data)
    
    # Calculate MSE for this fold
    mse_values[i] <- mean((test_data[[response]] - predictions)^2)
  }
  
  # Return results
  list(
    fold_mse = mse_values,
    mean_mse = mean(mse_values),
    sd_mse = sd(mse_values)
  )
}

df_values <- 1:10
results <- lapply(df_values, function(df) {
  cv_result <- cv_natural_spline(data = house_prices, "sale_price", "building_square_feet", df = df)
  c(df = df, mean_mse = cv_result$mean_mse)
})

# Convert results to data frame
results_df <- data.frame(do.call(rbind, results))

# Find best df
best_df <- results_df$df[which.min(results_df$mean_mse)]

# Fit final model with best df
final_formula <- as.formula(paste("sale_price ~ ns(building_square_feet, df =", best_df, ")"))
(final_model <- lm(final_formula, data = house_prices))

# Plot results
plot(house_prices$building_square_feet, log(house_prices$sale_price), main = "Natural Spline Fit with Best df", 
     xlab = "building_square_feet", ylab = "log(sale_price)", pch = 16, col = "gray")
x_new <- seq(min(house_prices$building_square_feet), max(house_prices$building_square_feet), length.out = 100)
pred <- log(predict(final_model, newdata = data.frame(building_square_feet = x_new)))
lines(x_new, pred, col = "red", lwd = 2)

```


# Compare model performance

```{r model-performance}
### AICc for glmnet models
### https://stackoverflow.com/questions/63171921/is-there-a-way-in-r-to-determine-aic-from-cv-glmnet
glmnet_cv_aicc <- function(fit, lambda = 'lambda.min'){
  whlm <- which(fit$lambda == fit[[lambda]])
  with(fit$glmnet.fit,
       {
         tLL <- nulldev - nulldev * (1 - dev.ratio)[whlm]
         k <- df[whlm]
         n <- nobs
         return(list('AICc' = - tLL + 2 * k + 2 * k * (k + 1) / (n - k - 1),
                     'BIC' = log(n) * k - tLL))
       })
}

```


| Model                        | Formula                                                                                                                      | AICc                              |
|------------------------------|------------------------------------------------------------------------------------------------------------------------------|-----------------------------------|
| Baseline                     | sale_price ~ building_square_feet                                                                                            | `r AICc(baseline)`                |
| Null                         | sale_price ~ 1                                                                                                               | `r AICc(null_model)`              |
| Full                         | sale_price ~ .                                                                                                               | `r AICc(full_model)`              |
| Step fit (forward selection) | sale_price ~ estimate_building + town_and_neighborhood + building_square_feet +	multi_property_indicator + property_class | `r AICc(step_fit)`                |
| Lasso                        |                                                                                                                              | `r glmnet_cv_aicc(lasso.cv)$AICc` |
| Ridge                        |                                                                                                                              | `r glmnet_cv_aicc(ridge.cv)$AICc` |
| Spline                       | sale_price ~ ns(building_square_feet, df = 5)                                                                                | `r AICc(final_model)`             |
